---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
df <-  read.csv("activity.csv")
df$date <- as.Date(df$date, format="%Y-%m-%d")
summary(df)
```

## What is mean total number of steps taken per day?
```{r}
library(dplyr)
x <- aggregate(df$steps, by=list(df$date), FUN=sum, na.rm=TRUE)
hist(x$x, xlab = "steps per day", main="")
cat("mean =", round(mean(x$x), digits = 0), 
    "\nmedian =",  median(x$x))
```


## What is the average daily activity pattern?
```{r}
int <- aggregate(df$steps, by=list(df$interval), FUN=mean, na.rm=TRUE)
plot(x = int$Group.1, y = int$x, type = "l", xlab = "5-min intermals", ylab = "steps")
cat("Max interval:", int[int$x==max(int$x),]$Group.1)
```



## Imputing missing values
```{r}
sum(is.na(df))
new_df <- df
# replace with interval average
for(i in 1:nrow(new_df)){
    if (is.na(new_df$steps[i]) == TRUE) {
      new_df[i,]$steps <- int[int$Group.1 == new_df[i,]$interval,]$x
    }}
# new hist
new_x <- aggregate(new_df$steps, by=list(new_df$date), FUN=sum, na.rm=TRUE)
hist(new_x$x, xlab = "steps per day", main="")
cat("mean =", round(mean(new_x$x), digits = 0), 
    "\nmedian =",  median(new_x$x))
```

These values differ from the estimates from the first part of the assignment. Mean got higher and now equals the median. Median stayed the same.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
library(timeDate)
new_df$day <- ifelse(isWeekday(new_df$date, wday=1:5), "weekday", "weekend")
new_df$day <- factor(new_df$day)

weekday_x <- aggregate(steps ~ interval, data=new_df[new_df$day=="weekday",], FUN=mean)
weekend_x <- aggregate(steps ~ interval, data=new_df[new_df$day=="weekend",], FUN=mean)
par(mfrow=c(2,1))
plot(x = weekday_x$interval, y = weekday_x$steps, type = "l", xlab = "5-min intermals", ylab = "steps", main="weekday")
plot(x = weekend_x$interval, y = weekend_x$steps, type = "l", xlab = "5-min intermals", ylab = "steps", main="weekend")
```

